---
title: "Cyclistic Bike Share Data Analysis"
author: "Jimi Adeleke"
date: "2023-09-30"
output: rmarkdown::github_document
---

## **Google Data Analytics Capstone Project: How Does a Bike-Share Navigate Speedy Success?**

### Introduction
The Google Professional Data Analytics Certification includes the Cyclistic case study. In this case study, I am a junior data analyst at Cyclistic. I will follow basic data analysis process steps to solve core business problems.

I'm a junior data analyst on the marketing analyst team at Cyclistic, a Chicago-based bike-sharing firm. The marketing manager believes that increasing annual memberships will be critical to the company's future success. As a member of the marketing team, he aims to discover how cyclistic bikes are used differently depending on membership type. My role here will be to study and analyze 12 months of data, extract insights from it using various visualizations, and present it to the team. I evaluated very precise data and attempted to substantiate these findings with clear graphs so that my proposals may be considered.In this case study, I used data from a fictional bike rental company based in Chicago, Cyclist, to try to determine and understand the main differences between annual members and casual rider. With such insights, I was able to deliver recommendations, with which the company has developed a marketing campaign aimed at converting casual riders to annual members, and hence maximizing the growth of the company.


I obtained the datasets from the link provided below.

Cyclistic travel data for 2022 is available at <https://divvy-tripdata.s3.amazonaws.com/index.html>.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up the environment

Here, I use several libraries that help reading, cleaning, organizing and analyzing the data.
```{r load-packages, include=FALSE}
library(tidyverse) # data manipulation and visulization
library(lubridate) # used for date format
library(here) # display the top level directory of the current project
library(hms) # a simple class for storing duration or time-of-day
library(data.table) # for fast aggregation of large datasets
library(janitor) # for examining and cleaning dirty data
library(skimr) # alternative to summary(), provide broad overview of a data frame.
library(hablar)# provide a dynamic way to change type of a column
library(readxl) # make it easy to get data out of excel and into r
library(chron) #Chronological object that can handle dates and time
library(readr) #provide a fast and friendly way to read rectangular data from delimited files
library(magrittr) #improve readability and maintainability of code
library(DescTools) # for statiscally describing data
library(metR) # for easier analysis of meteorological fields
```

#### Importing data 
Cyclist data from 01/2022 until 12/2022 is imported and read as csv. files.

```{r csv data collection}
data_1 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202201-divvy-tripdata.csv")
data_2 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202202-divvy-tripdata.csv")
data_3 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202203-divvy-tripdata.csv")
data_4 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202204-divvy-tripdata.csv")
data_5 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202205-divvy-tripdata.csv")
data_6 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202206-divvy-tripdata.csv")
data_7 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202207-divvy-tripdata.csv")
data_8 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202208-divvy-tripdata.csv")
data_9 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202209-divvy-tripdata.csv")
data_10 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202210-divvy-tripdata.csv")
data_11 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202211-divvy-tripdata.csv")
data_12 <- read.csv("C:/Users/USER/Desktop/PORTFOLIO PROJECT/R/Google data analysis capstone (bike share analysis)/Datasets/202212-divvy-tripdata.csv")
```

## checking datasets columns

Compare column names each of the files. While the names don't have to be in the same order, they DO need to match perfectly before they are  joined  into one file.
```{r}
colnames(data_1)
colnames(data_2)
colnames(data_3)
colnames(data_4)
colnames(data_5)
colnames(data_6)
colnames(data_7)
colnames(data_8)
colnames(data_9)
colnames(data_10)
colnames(data_11)
colnames(data_12)
```

## combining all datasets into 1
the 12 data sets are combined into one big data frame of 5667717 observations and 13 columns
```{r}
full_year <- rbind(data_1,data_2,data_3,data_4,data_5,data_6,data_7,data_8,
                   data_9,data_10,data_11,data_12)
dim(full_year)
```

##  Clean up and organize data to prepare for analysis
Inspect the new table that has been created.

#### see the first 6 rows
```{r}
head(full_year) 
```
####  List of column names
```{r}
colnames(full_year) 
```
####  list of columns and data types (numeric, character, etc)
```{r}
str(full_year) 
```
#### create columns for: day of week, month, day, year, time, hour 
days of the week are assigned the numbers 1:Monday, 2:Tuesday, etc.
This will allow the aggregation of the data by each day, month or day_of_week.
```{r}
full_year$date <- as.Date(full_year$started_at) #The default format is yyyy-mm-dd
full_year$year <- format(as.Date(full_year$date),'%Y') #extracting year from date
full_year$month <- format(as.Date(full_year$date),'%m') #extracting month from date
full_year$day <- format(as.Date(full_year$date),'%d') #extracting day from date
full_year$num_of_day <- format(as.Date(full_year$date),'%u') #extracting num_of_week from date
full_year$day_of_week <- format(as.Date(full_year$date),'%A') #extracting day_of_week from date
#full_year$time <- format(as.Date(full_year$started_at), "%H:%M:%S") #format time as HH:MM:SS
full_year$time <- format(as.POSIXct(full_year$started_at), format = "%H:%M:%S")
#full_year$hour <- format(as.POSIXct(full_year$started_at), format = "%H") #extracting hour from time
```
#### extracting hour from time
```{r}
full_year$hour <- hour(full_year$started_at)
```

#### create column for different seasons: Spring, Summer, Fall, Winter
```{r}
# full_year$season <- season(full_year$month) (use to assign season to month)
full_year <- full_year %>%
  mutate(season = 
           case_when(month == "03" ~ "Spring",
                     month == "04" ~ "Spring",
                     month == "05" ~ "Spring",
                     month == "06"  ~ "Summer",
                     month == "07"  ~ "Summer",
                     month == "08"  ~ "Summer",
                     month == "09" ~ "Fall",
                     month == "10" ~ "Fall",
                     month == "11" ~ "Fall",
                     month == "12" ~ "Winter",
                     month == "01" ~ "Winter",
                     month == "02" ~ "Winter"))
```
#### create column for different time_of_day: Night, Morning, Afternoon, Evening
```{r}
full_year <- full_year %>%
  mutate(time_of_day =
           case_when( hour == "00" ~ "Night",
                      hour == "01" ~ "Night",
                      hour == "02" ~ "Night",
                      hour == "03" ~ "Night",
                      hour == "04" ~ "Night",
                      hour == "05" ~ "Night",
                      hour == "06" ~ "Morning",
                      hour == "07" ~ "Morning",
                      hour == "08" ~ "Morning",
                      hour == "09" ~ "Morning",
                      hour == "10" ~ "Morning",
                      hour == "11" ~ "Morning",
                      hour == "12" ~ "Afternoon",
                      hour == "13" ~ "Afternoon",
                      hour == "14" ~ "Afternoon",
                      hour == "15" ~ "Afternoon",
                      hour == "16" ~ "Afternoon",
                      hour == "17" ~ "Afternoon",
                      hour == "18" ~ "Evening",
                      hour == "19" ~ "Evening",
                      hour == "20" ~ "Evening",
                      hour == "21" ~ "Evening",
                      hour == "22" ~ "Evening",
                      hour == "23" ~ "Evening"))
```
#### create a column for the month using the full month name
```{r}
full_year <- full_year %>%
  mutate(month_name = 
           case_when( month == "01" ~ "January",
                      month == "02" ~ "February",
                      month == "03" ~ "March",
                      month == "04" ~ "April",
                      month == "05" ~ "May",
                      month == "06" ~ "June",
                      month == "07" ~ "July",
                      month == "08" ~ "August",
                      month == "09" ~ "September",
                      month == "10" ~ "October",
                      month == "11" ~ "November",
                      month == "12" ~ "December"))
```
#### calculate ride length by subtracting ended_at time from started_at time and converted it to minutes
```{r}
full_year$ride_length <- difftime(full_year$ended_at, full_year$started_at, units='mins')
full_year$ride_length <- round(full_year$ride_length, digits =1)
```

#### Convert c(ride_length, ride_length_m, day and month) to numeric so that calculation can be executed.
```{r converting variables to numeric, echo=TRUE, eval=TRUE}
full_year$ride_length <- as.numeric(as.character(full_year$ride_length))
#all_trips$ride_length_m <- as.numeric(as.character(all_trips$ride_length_m))
full_year$month <- as.numeric(full_year$month)
full_year$day <- as.numeric(full_year$day)
is.numeric(full_year$ride_length)
#is.numeric(full_year$ride_length_m)
is.numeric(full_year$month)
is.numeric(full_year$day)
```
#### checking the dtypes of each column
```{r}
str(full_year)
```
#### checking ride_length column if there is negative number
```{r}
#full_year %>%
  #filter(ride_length < 0)
```
#### remove negative number from ride_length column
```{r}
full_year <- full_year[!(full_year$ride_length <= 0),]
```
#### removing rows with NA
```{r}
full_year <- na.omit(full_year)
```
#### remove duplicate rows
```{r}
full_year <- distinct(full_year)
```
#### remove columns not needed:ride_id, start_station_id, end_station_id, start_lat, start_lng,end_lat,end_lng
```{r}
full_year <- full_year %>%
  select(-c(ride_id, start_station_id, end_station_id, start_lat, start_lng,end_lat,end_lng))
```
#### checking the num of obs and variables of the cleaned dataset
```{r}
dim(full_year)
```
The new dataset has 4572403 observations and 18 variables
#### view the clean dataset
```{r}
head(full_year)
```
## Decriptive Data Analysis
#### noumber of rows
```{r}
nrow(full_year)
```
In this data frame, we have a total of 4,572,403 rides. It took place between January 2022 and December 2022.

### Total Rides by User Type
```{r}
full_year %>%
  group_by(member_casual) %>%
  summarise(total_rides = n())

full_year %>%
  group_by(member_casual) %>%
  summarise(total_rides = n()) %>%
  ggplot(aes(x=member_casual, y=total_rides, label = total_rides)) +
      geom_bar(position = "dodge", stat = 'identity') +
      geom_text(size = 5, position = position_dodge(width = .9), vjust=0)+
      labs(title = 'Total Rides by User Type', x='user_type', y='total_rides')
```
The total rides of casual is 1,965,432 and that of member is 2,606,971
### Total rides by week_day

```{r}
full_year %>% 
  group_by(day_of_week) %>% 
  summarise(number_of_rides = n())

full_year %>% 
  group_by(day_of_week) %>% 
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = reorder(day_of_week,-number_of_rides), y = number_of_rides, label = number_of_rides)) + 
  geom_bar(position = 'dodge', stat = 'identity')+
  geom_text(size = 4, position = position_dodge(width = .9), vjust =-0.5) +
  labs(title = 'Total rides by week_day', x='day_of_week', y = 'number_of_rides')
```
When we examine according to the days of the week, it has been observed that most weekend driving takes place. It is assumed that the most important reason for this is that the members prefer to tour by bike as a weekend activity, since they have weekend holidays.

### Riding weekday by the Member Type
```{r}
full_year %>% 
  group_by(day_of_week, member_casual) %>% 
  summarise(number_of_rides = n(), .groups = 'drop') %>%
  ggplot(aes(x = day_of_week, y=number_of_rides, fill = member_casual)) + 
  geom_bar(position = 'dodge', stat = 'identity')+
  #geom_text(size = 4, position = position_dodge(width = .9), vjust =-0.5) +
  labs(title = 'Riding weekday by the Member Type', x='day_of_week', y = 'number_of_rides')
```
If we divide this analysis according to member types, it is observed that casual members reach the highest number of rides on Saturdays, while annual members reach the highest number of rides on Thursdays. Annual members made more rides on all other days in total number of rides except Saturday.

### Total Rides by per Bike Type
```{r}
full_year %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(number_of_rides = n(), .groups = 'drop')

full_year %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(number_of_rides = n(), .groups = 'drop') %>%
  ggplot(aes(x = rideable_type, y = number_of_rides, label = number_of_rides, fill=member_casual)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  geom_text(size = 3.5, position = position_dodge(width = .9), vjust = 0) +
  labs(title = 'Total Rides by per Bike Type', x = 'number of rides', y= 'rideable type')
```
It was observed that casual members preferred the most electric bike type with 1,041,016 rides, while annual members preferred the classic bike type the most with 1,336,588 rides. 

### Riding by the Hours
```{r}
full_year %>%
  group_by(member_casual, hour) %>%
  summarise(total_ride = n(), .groups = 'drop') %>%
  ggplot(aes(x = hour, y = total_ride, color = member_casual)) +
  geom_line() +
  scale_x_continuous(breaks = seq(4,24, by=2))

full_year %>%
  group_by(member_casual, hour) %>%
  summarise(total_ride = n(), .groups = 'drop') %>%
  ggplot(aes(x = hour, y = total_ride, fill = member_casual)) +
  geom_bar(stat = 'identity')
```
When the number of driving hours of the day according to the user type is examined, it is clearly seen that the annual and casual members drive more especially in the afternoon and evening hours. 

### Riding by time of the day
```{r}
full_year %>%
  group_by(member_casual, time_of_day)%>%
  summarise(total_rides = n(), .groups = 'drop') %>%
  ggplot(aes(x = time_of_day, y = total_rides, fill = member_casual, label=total_rides)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  geom_text(size = 3, position = position_dodge(width = .9), vjust=-0.5) +
  labs(title = 'riding by time_of_day', x = 'time_of_day', y='total_rides')
```
The driving behaviors of the users in certain time periods of the day were examined. According to these reviews, it has been observed that annual and casual members have a higher driving percentage in the Afternoon compared to other time periods. While the evening time period has the second highest percentage of driving, the least driving occurred at morning.

### number of rides by months
```{r}
full_year %>%
  group_by(member_casual, month_name) %>%
  summarise(total_rides = n(), .groups = 'drop') %>%
  ggplot(aes(x = reorder(month_name, -total_rides),  y = total_rides, fill=member_casual)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  #geom_text(size = 3, position = position_dodge(width = .9), vjust=0) +
  labs(title = 'number of rides by months', x = 'month', y='number of rides') +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust =1))
```
When we group our total number of rides by months, we observe that the highest number of rides occurs in july. Again, August and June were observed as the months with the most driving. It was clearly seen that December, January and February had the lowest driving percentages. 
### number of rides per season
```{r}
full_year %>%
  group_by(member_casual, season) %>%
  summarise(total_rides = n(), .groups = 'drop') %>%
  ggplot(aes(x = season, y = total_rides, label = total_rides, fill=member_casual)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  geom_text(size = 3, position = position_dodge(width = .9), vjust=0) +
  labs(title = 'number of rides per season', x = 'season', y='number of rides')
  #theme(axis.text.x = element_text(angle = 45, vjust=1, hjust =1))
```
Both the casual and member user_type drive most in the summer season. 
```{r}
full_year%>%
  group_by(season, day_of_week, member_casual) %>%   
  summarise(number_of_rides = n()						 
  ,avg_ride_length = mean(ride_length), .groups = 'drop') %>% 
  ggplot(aes(x = day_of_week, y = avg_ride_length,fill=member_casual)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  #geom_text(size = 3, position = position_dodge(width = .9), vjust=0) +
  facet_wrap(~season) + 
  scale_y_continuous(breaks = seq(0, 50, by = 10)) +
  labs(title = 'ride_lenght by week_day', x = 'day_of_week', y='avg_ride_length') +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust =1))
#ggplot() + geom_bar(mapping = aes(x = day_of_week, y = avg_ride_length, fill = member_casual), #position = "dodge") + facet_wrap(~season) + scale_y_continuous(breaks = seq(0, 50, by = 10))
```
demonstrates that the member group has all year long the average ride length of about 13.6 minutes. Casual riders use bikes  about half an hour long on all days in spring and summer. In winter and fall, the average ride lengths becomes less than 30 minutes.
```{r}
full_year%>%
  group_by(season, day_of_week, member_casual) %>%   
  summarise(number_of_rides = n()						 
  ,avg_ride_length = mean(ride_length), .groups = 'drop') %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill=member_casual)) +
  geom_bar(position = 'dodge', stat = 'identity') +
  #geom_text(size = 3, position = position_dodge(width = .9), vjust=0) +
  facet_wrap(~season) + 
  scale_y_continuous(breaks = seq(0, 400000, by = 50000)) +
  labs(title = 'number of rides by week_day', x = 'day_of_week', y='number_of_rides') +
  theme(axis.text.x = element_text(angle = 45, vjust=1, hjust =1))
#ggplot() + geom_col(mapping = aes(x = day_of_week, y = number_of_rides, fill = member_casual), #position = "dodge") + facet_wrap(~season) + scale_y_continuous(breaks = seq(0, 400000, by = 50000))
```
tells us that number of rides of members is always higher than that of casual riders on every  work day in every season. Weekends are still the time where casual riders bike more than members. The only exception of this trend is in the winter months (Dec, Jan, Feb).
```{r}
full_year %>%
  group_by(member_casual, month) %>%
  summarise(total_ride = n(),
            avg_ride_length = mean(ride_length), .groups = 'drop') %>%
  ggplot(aes(x = month, y = total_ride, color = member_casual)) +
  geom_line() +
  scale_x_continuous(breaks = seq(1,12, by=2))+
  labs(title = 'number of rides along the whole year')

```
The plot indicates, for casual riders/members, ridership peaked around July/August (Summer months being the turning point) and hit the lowest at February before rebounding up swiftly and continuously.




